---
title: "Using dplyr for data handling"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output: html_document
---

For these exercises, we will use a new dataset related to mammalian sleep as provided in Data Analysis for the Life Sciences (Rafael Irizarry, Mike Love).
 
 We are going to read in this data, then test your knowledge of they key `dplyr` functions. We are also going to review two different *classes*: data frames and vectors.
 
###  0 - Load necessary libraries
 0. Load the dplyr library or the tidyverse collection of libraries that includes dplyr.
```{r}
library(tidyverse)
```

###  1 - Read in and inspect the data 
1a. Read in the `msleep_ggplot2.csv` file with the function `read.csv` into an object called `msleep`.
```{r}
msleep <- read.csv("~/postdoc/teaching/RNAseqCourse/ExerciseData/msleep_ggplot2.csv")
```

1b. Use the function `class` to determine what type of object is `msleep`.
```{r}
class(msleep)
```

1c. Use the `head` function or `View()` to take a glimpse at the data.
```{r}
head(msleep)
```

1d. How many animals are in the data in total and how many are there for each order? Hint: Use `nrow` and `table`
```{r}
nrow(msleep)
```


###  2 - Modify the data
2. Add a new column to the data that contains the fraction of REM sleep of the total sleep using the `mutate` function.
```{r}
msleep <- msleep %>% mutate(REM_fraction = sleep_rem / sleep_total)
```

###  3 - Subset the data to primates
3a. Now use the `filter` function to select only the primates from the `msleep` object and give the resulting object a new name.
```{r}
msleep_primates <- filter(msleep, order == "Primates")
nrow(msleep_primates)
```

3b. How many primates species are in the data?
```{r}
nrow(msleep_primates)
```

3c. What is the class of the object you obtain after subsetting the table to only include primates?
```{r}
class(msleep_primates)
```

###  4 - Calculate the average sleep of primates
4a.  Now use the `select` function to extract the sleep (total) for the primates. 
```{r}
sleep_total_primates <- select(msleep_primates, sleep_total)
```

4b. What class is this object? 
```{r}

class(sleep_total_primates)
```

4c. Calculate the average amount of sleep for primates using the `mean` function.
```{r}
mean(sleep_total_primates$sleep_total)
```

Alternative with pipe
```{r}
filter(msleep, order == "Primates") %>%
    .$sleep_total %>%
    mean()
```

###  5- Advanced exercises
To calculate the average amount of sleep for primates, we could also use the dplyr `summarize` function. We have not introduced this function. For unkown functions it is useful to look at the R help providing details on a function.

5a. Use `?summarize` to open the help page for the `summarize` function. 

5b. Calculate the average amount of sleep for primates starting from the `msleep` object in a single code line using `filter`, the pipe `%>%` and `summarize` to get the answer.
```{r}
# to be filled
msleep %>%
    filter(order == "Primates") %>%
    summarise(mean_total_sleep = mean(sleep_total))
```

The `summarize` function can be combined with the `group_by` function to calculate a summary statistics (e.g., the mean) for different groups of rows in a data frame. For example, this allows us to calculate the average amount of sleep for all orders of mammals in the data without having to repeat the exercises 3-6 for all orders separately. 

5c. Take a look at the help of `?group_by`.

5d. Group the data frame obtained in exercise 1 (`msleep`) by "order" and use `summarise` on the grouped data frame to calculate the average amount of sleep for each order.
```{r}
msleep %>%
    group_by(order) %>%
    summarise(mean_total_sleep = mean(sleep_total))
```

In addition to the mean we can compute other summaries, for example the function `n()` counts the number of rows in each group. 

5e. Use `group_by` to group the data frame obtained in exercise 1 by "order" and `summarise` the grouped data frame to generate a new data frame with the average total sleep and number of species per order.
```{r}
msleep_by_order <- msleep %>% 
  group_by(order) %>%
  summarise(mean_total_sleep = mean(sleep_total),
            number_species = n())
msleep_by_order
```

5f. Use the data frame obtained in exercise 5e and the function `arrange` to sort the rows of the data.frame by the average amount of sleep. Which mammalian order sleeps the most?
```{r}
arrange(msleep_by_order, -mean_total_sleep)
```

### Final words
6. Call the function `sessionInfo()` to have an output of your R and package versions and operating system. This is good practice to always include so that other people can reproduce your code.
```{r}
sessionInfo()
```

7. Knit your file to produce a html report on your analysis
