---
title: RNAseq Normalization 
author: "Nikolai Köhler"
date: "`r Sys.Date()`"
output: html_document
---

```{r i_question, include = FALSE, echo = FALSE}
qi <- 0
```

# Importing data and loading required packages

In this exercise we are using a simulated dataset from the `DESeq2` package.
We will be talking more about its capabilities on Friday.

The below code simulates a dataset with 40 samples and 2000 genes

```{r md_setup, message=FALSE}
if (!requireNamespace("DESeq2", quietly=TRUE))
    BiocManager::install("DESeq2")

library(DESeq2)

# we set a random seed to make sure results are consistent across sessions
set.seed(42)

de_data <- makeExampleDESeqDataSet(n=2000, m=40)
de_data
```

# Normalizing to Sequencing Depth and Gene Length {#simple-norm}

The simplest form of normalization accounts for sequencing depth and gene length.
Lucky for us, these methods are quite simple to implement.

## Counts per Million Reads Mapped

Compute the counts per million reads mapped (CPM). 
To obtain the raw counts from the `de_data` object, you can use the `counts` function.

```{r cpm}
cpm_corrected <- # TODO
head(cpm_corrected[,1:10])
```

**Question `r (qi <- qi + 1)`**: What does CPM control for? Is it suited for within-sample comparisons?


## FPKM

Next, we compute the FPKM value. To get the genomic ranges, you can use `rowRanges`.
`width` will give you the 

```{r fpkm}
fpkm_corrected <- # TODO
head(fpkm_corrected[,1:10])
```


## FPKM in `DESeq2`

Get familiar with `DESeq2`'s `fpkm` function and run it below.

```{r deseq}
# TODO
```

# Advanced Normalisation Techniques

In the lecture we talked about methods such as variance stabilizing transformation (VST) and trimmed mean of M values (TMM).

**Question `r (qi <- qi + 1)`**: What is the overarching idea of these techniques as compared to the methods in [the previous section](#simple-norm)?

*Important*: Neither of these techniques is ideal for testing differential gene expression!
We will talk about how to do this properly on Friday.

Unfortunately, these techniques are a little too complicated to be implemented in this exercise. Therefore, we'll resort to an implementation available in `DESeq2`. We will dive deeper into `DESeq2` on Friday.
For now, it is sufficient to only look at the functions mentioned below.


## Variance Stabilizing Transformation

In this section we are using DESeq2 to perform variance stabilizing transformation (VST).
VST in implemented in `getVarianceStabilizedData`.
To compute it, size and dispersion factors need to be computed beforehand.
The corresponding functions are `estimateSizeFactors` and `estimateDispersions`.

```{r}
de_data <- # TODO: factor size and dispersion estimation
vsd <- # TODO: variance stabilized data
```

Plot two histograms comparing the raw counts with the VST data.
```{r}
hist(counts(de_data))
```

```{r}
hist(vsd)
```

Another way of visualizing the effect of VST is by plotting the rank of means against the standard deviation, for example using the `meanSdPlot` function from the vsn package.

Use it to show the difference between raw counts, log2 transformed data and VST transformed data.
*Hint*: use `cowplot::plot_grid` to plot all three into one figure.

```{r}
library(vsn)

cowplot::plot_grid(
    plotlist=list(
        meanSdPlot(counts(de_data))$gg,
        meanSdPlot(log2(counts(de_data) + 1))$gg,
        meanSdPlot(vsd)$gg
    )
)
```

## Advanced (if we have time)

Implement your own version of quantile normalization.

Recall the steps for quantile normalization:

1. Order genes by expression for each sample
2. Compute mean expression *over all samples* for *each rank*
3. Assign mean values to samples
4. Reorder back to original gene ordering

If you fell very confident (i.e. *even more advanced*): define a function taking in a data.frame and returning its quantile normalized data.
