---
title: "PCA on the CLL expression data"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output: html_document
---

# Preparations

0a. Load the following package.
```{r}
library(tidyverse)
```

0b. Install and load the ggfortify package.
```{r}
# to be filled
```

# Load and subset the data
For this exercise we will use two data sets from the CLL data, containing different molecular measurements from chronic lymphocytic leukemia patient samples. We will make use of:

- gene expression data `mRNA` slot: This data contains normalized gene expression measurements for 5,000 genes

- additional patient information on the Gender and the so-called IGHV status (clinical marker used for diagnosis and patient stratification) of the patient.

1a. Load the expression data from the MOFAdata package that can found in the ExerciseData.zip in Heibox and subset to the patient samples with complete measurements by running the following code and inspect its output.
```{r}
load("put_your_path_here/MOFAdata.RData")

# obtain RNA data for patients with complete RNA measurements
mRNA <- CLL_data$mRNA # get mRNA data matrix from the CLL data list
mRNA <- t(mRNA) # transpose matrix to have samples in rows
has_no_mRNA <- apply(mRNA, 1, function(x) !any(is.na(x))) # find patient samples with complete mRNA information
complete_patients <- names(has_no_mRNA)[has_no_mRNA]
mRNA_complete <- mRNA[complete_patients,]

# obtain information on Gender and IGHV status for those patients and collect them in a dataframe
IGHV <- CLL_data$Mutations["IGHV", complete_patients]
trisomy12 <- CLL_data$Mutations["trisomy12", complete_patients]
gender <- CLL_covariates[complete_patients, "Gender"]
df_patinfo <- data.frame(patient_sample = complete_patients, 
                         IGHV = as.factor(IGHV),
                         gender = gender,
                         trisomy12 = as.factor(trisomy12))
```

1b. How many samples and features are in the `mRNA_complete` object?

# PCA
2 Let's calculate the PCA for the gene expression measurements.
2a. Use the `prcomp` function on the `mRNA_complete` object to calculate a PCA.
```{r}
# to be filled
```

2b. Use the `autoplot` function on your results to plot the PCA for component 1 versus 2 as well as 3 versus 4. Use the `coord_fixed()` option for all plots (what does it do?). 
```{r}
# to be filled
```

2c. Color the plots from 2b first by Gender, then by IGHV and then by trisomy 12.  
```{r}
# to be filled
```

2d. Look at the `rotation` matrix in the PCA object to find which genes contribute most to the first prinicipal component.
```{r}
# to be filled
```

2e. Use the `summary` function on the PCA object to get information on the variance explained by each component
```{r}
# to be filled
```

2f. Extract the variance explained and make an Elbow plot to determine a good number of PCs to keep.
```{r}
# to be filled
```

# PCA on other omic layers
3. Optional: Compare the structure between patient samples that you found in the PCA plot from gene expression data to the structure in the drug response data.
3a. How does a PCA of the patient on the drug response data looks like? 
3b. How does IGHV or tisomy12 relate to it?
```{r}
# to be filled
```

```{r}
sessionInfo()
```

