---
title: "Differential expression between male and female patient samples"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output: html_document
---

Here we will test for differential expression of genes in the CLL data. Note that dedicated methods exists for the task of differential expression testing that make use of (unnormalized) counts for each gene and patient sample instead of normalized expression values. These can be much more powerful and better calibrated for RNA-seq data and we will learn more about them in the summer course.

# Preparations

0. Load the ggplot2 and dplyr library or the tidyverse collection of libraries that includes ggplot2 and dplyr.
```{r}
# to be filled
```

For these exercises, we will use again the CLL data from the MOFAdata package that contains different molecular read-outs on a set of patient samples.
```{r}
library(MOFAdata)
data("CLL_data")
data("CLL_covariates")
CLL_covariates <- rownames_to_column(CLL_covariates, "patient_sample")

# (alternative) You can also load from the folder ExerciseData:
load("ExerciseData/MOFAdata.RData")
CLL_covariates <- rownames_to_column(CLL_covariates, "patient_sample")
```


# Part I: Differential expression testing

1. Create a dataframe with gene expression and sex information following similar steps as in the ggplot exercise on this data to remove patients with no RNA measurements and match the gene expression and the patient information using for example `left_join`. Verify that you have 136 patients in your final data.

```{r}
# to be filled
```

2. Select one gene and conduct a t-test of its expression values for male and female samples. How can you extract the p-value from the test results? Take a look at the help of ?t.test to find out. Then plot a boxplot of the genes expression for male and female samples.
```{r}
# to be filled
```

3. Repeat step 2 to compute the p-values for all genes using a for-loop or the `sapply` function.
```{r}
# to be filled
```

4. How many p-values are below 0.05?
```{r}
# to be filled
```

5. Plot a boxplot for some of the genes with a p-value < 0.05.
```{r}
# to be filled
```

6. How many genes have a p-value below the threshold for a family wise error rate of 0.05 (remember this is 0.05 / total number of tests)? Plot a boxplot of their expression values for male and female samples
```{r}
# to be filled
```

7. Use `p.adjust` function with the Bonferroni method to control the family-wise error rate at 5%. How many values are below 0.05 after adjustement?
```{r}
# to be filled
```

8. Use `p.adjust` function with the Benjamini-Hochberg method to control the false discovery rate (FDR) at 5%. How many genes are significant at a FDR of 5%?
```{r}
# to be filled
```

9. We can visualize the results of the testing for all genes in a so-called volcano plot that shows the effect size (difference in mean gene expression between male and females) and significance (e.g. negative adjusted p-value on a log-scale) for all genes. For this calculate the mean expression difference for each gene (using a for loop or sapply) and create a data.frame containing for each gene the gene name, BH-adjusted p-value and mean difference. You can also add a column to the data.frame specifying whether a gene is significant at a FDR level of 5%.
```{r}
# to be filled
```

10. Use this data frame to make a volcano plot that shows the mean difference on the x-axis and `-log( BH-adjusted p-value)` on the y-axis using `geom_point`. How many genes stand out? You can also color the points by whether or not a gene is significant. Look up some of the genes that stand out on ensemble.org - what do you find?
```{r}
# to be filled
```

```{r}
sessionInfo()
```

