---
title: "biomart Introduction"
author: "Nikolai Köhler"
date: "2025-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!require("biomaRt", quietly = TRUE))
    BiocManager::install("biomaRt")

library(biomaRt)
library(tidyverse)
```

## Load ensembl mart


Load ensembl mart for **human genes**

**Task 1**: setup a mart object containing information on genes for the 
human dataset using the `useEnsembl` function.


*Hints*:
* `listMarts` gives a list of all available marts
* `listDatasets` gives a list of available datasets for a given mart

```{r mart}
ensembl <- # TODO
```

## Query gene information

*Hints*:

* to list available attributes use `listAttributes`, to search use `searchAttributes` (also allows search by pattern)
* to find available filters use `listFilters`
* to retrieve information use `getBM` with the appropriate `attributes, `filters`, `values`, and `mart` parameter set

```{r listAttrs}
# TODO
```

```{r listFilts}
# TODO
```

**Tasks**

Given the following Ensembl ID **ENSG00000141510**, find ...

...the gene name (HGNC symbol) and description
```{r get-name}
# TODO
```

...all associated transcripts
```{r get-transcripts}
# TODO
```

...the chromosome and the position of the gene on the chromosome
```{r get-position}
# TODO
```

# Multi-gene queries

## Finding gene IDs from names

For the following genes, retrieve the corresponding Ensembl gene IDs: BRCA1, MYC, EGFR, PTEN

*Hint*: check how to pass multiple genes to `values`

```{r get-ids}
target_genes <- c("BRCA1","MYC","EGFR","PTEN")
# TODO
```

**Task**: count the number of transcripts per gene
```{r count-transcripts}
# TODO
```


In addition to genetic information, ensembl also contains functional annotations,
such as Gene Ontology terms.

**Task** for the genes given above, retrieve the GO annotations and find the
GO terms that appear most frequently. 

```{r get-go}
go_counts <- # TODO
head(go_counts)
```
Side notes:

* this also works the other way round - you can find all genes annotated with a particular GO term)
* you can retrieve information on GO terms with, for example, the `GO.db` package

**Task**: plot a histogram of the number of occurences of the go terms

```{r}
# TODO
```

# Finding Analogs

`biomaRt` also allows to find links between datasets, typically used for finding information
across organism. The associated `R` function is `getLDS` (get linked dataset).

Use `getLDS?` to find the parameters you need to pass to the function and find the chromosome
and start position of ENSG00000141510/TP53 on the mouse genome.

```{r get-analog}
# TODO
```

```{r session-info}
sessionInfo()
```